INGRID: Creating Languages in MPS from ANTLR Grammars

### 2016-10-24 chapters/language_definition.tex
parsing

The process of MPS language construction, as performed by INGRID, consists of four steps --- parsing of the input grammar, followed by definition of the essential aspects (Structure, Editor, and TextGen).
Each of the steps 2-4 is exclusively and fully responsible for one aspect.

@@

The grammar of SimpleXML contains elements of the kinds that are listed below.
Each color in Figure~\ref{fig:SIMPLEXML} corresponds to a different kind in a way that is indicated by the list item headers.
\begin{itemize}
	\item \textbf{ANTLR v4 keywords} are required by the notation.
	\item \antlrparserrule{\textbf{Parser rules}} describe the structure of a language.
		Alternatives on the right side of a rule are separated by the pipe character ($|$).
	\item \antlrlexerrule{\textbf{Lexer rules}} describe terminal symbols that the parser matches against the input string.
		A terminal symbol can be encoded as a string value or using a regular expression.
		The \textbf{fragment} keyword states that the lexer rule just brings more clarity into the notation but it is not visible in the parser output.
	\item \antlrliteralnoap{\textbf{String literals}}, always written inside of a pair of single quote marks, also represent terminal symbols, but by exact match to a string constant.
	\item \antlrregex{\textbf{Regular expressions}} describe string tokens to be matched against regular expressions with a special ANTLR v4 regex notation\footnote{https://github.com/antlr/antlr4/blob/master/doc/lexer-rules.md}.
\end{itemize}
Elements on the right side of a rule can be annotated with standard EBNF operators (?+*) that specify the allowed number of occurrences.

@@

\subsection{Parsing (the) Grammar}

\todo{Define some name for the AST representing the grammar, and use it consistently throughout the whole paper. Maybe distinguish between "grammar AST" and "MPS structure AST".}
The first step (stage) of creating the MPS language (of the import process) is parsing the grammar file (written using the ANTLRv4 notation) so that we can get an AST tree representing the grammar.

The abstract syntax tree that comes out of the automatically generated (ANTLR) parser is a little bit complicated and full of information not relevant for our purposes (to us).
That is why we translate the complex ANTLR AST into our own simplified tree made out of our own objects.
We discarded as much irrelevant information as possible and only kept information (data) vital/necessary for construction of MPS languages.

Furthermore, we process the grammar in two iterations.
In the first iteration/pass, we build a tree holding names of other referenced rules in the form of strings.
In the second iteration, we walk this tree and resolve all these references to real pointers to other rule objects.
We end up with a clear and simpler representation that is easy to process in the next steps of the import process.

We also had to post-process (simplify) the representation of lexer rules (tokens).
As a result of the initial parsing, ANTLR parser produces some kind of a tree structure that is representing the lexer rule (meaning/note that lexer rules can also be built from alternatives just like parser rules).
However, lexer rules are, in fact (basically), just regular expressions used for parsing input source code into tokens.
We took each/every such tree that represents the lexer rule and flattened it into the equivalent regular expression.

For illustration, consider the lexer rule \antlrlexerrule{Name} from our/the SimpleXML language (Figure XY), together with all its sub-rules, shown in Figure~\ref{fig:lexerrule}.

\begin{figure}[ht]
\centering
\begin{framed}
\begin{alltt}
	\antlrlexerrule{Name}     : \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
	\textbf{fragment}
	\antlrlexerrule{DIGIT}    : \antlrregex{[0-9]} ;
	\textbf{fragment}
	\antlrlexerrule{NameChar} : \antlrlexerrule{NameStartChar} 
	         | \antlrliteral{-} | \antlrliteral{{\_}} | \antlrliteral{.} | \antlrlexerrule{DIGIT} ;
	\textbf{fragment}
	\antlrlexerrule{NameStartChar} : \antlrregex{[:a-zA-Z]} ;
\end{alltt}
\end{framed}
\caption{Lexer rule from SimpleXML}
\label{fig:lexerrule}
\end{figure}

Intuitively, we can create a regular expression by gluing elements of each alternative together and then join these alternatives with a classic regex OR ($|$), which is exactly its function in the ANTLRv4 grammar notation.
We have captures this idea in the form of the recursive algorithm in Figure~\ref{fig:algflatten}, which covers the whole process.

% TODO prepsat jako standardni pseudocode (nejspis pouzit jiny latex environment)
\begin{figure}[ht]
\centering
\begin{framed}
\begin{alltt}
Flatten(\textit{R}):
  Define \textit{T} as a list of empty string
  For each alternative \textit{A} of the rule \textit{R}:
    Define \textit{R} = \antlrap\antlrap
    For each element \textit{E} of \textit{A}:
	  If \textit{E} is not yet flattened:
	    Flatten(\textit{E})
      \textit{R}.append(\textit{E})
      \textit{R}.append(\textit{E}.operator)
    \textit{T}.add(\textit{R})
  Build a string S from elements of \textit{T}:
    (t$_{\text{1}}$)|(t$_{\text{2}}$)|...|(t$_{\text{n}}$)
  Return \textit{S}
\end{alltt}
\end{framed}
\caption{Flatten algorithm}
\label{fig:algflatten}
\end{figure}

The output of \texttt{Flatten(Name)} (output of the Flatten algorithm applied to the \antlrlexerrule{Name} rule) is this regular expression (in this form):

\begin{center}
	\texttt{[:a-zA-Z]([:a-zA-Z]|\textbackslash-|{\_}|\textbackslash.|[0-9])*}
\end{center}

>>>>

The process of MPS language construction, as performed by INGRID, consists of four phases --- the first is parsing of the input grammar, which is followed by definition of the essential aspects (Structure, Editor, and TextGen).
Each of the phases 2-4 is exclusively and fully responsible for one aspect.

@@

The grammar of SimpleXML contains elements of the kinds that are listed below.
Each color in Figure~\ref{fig:SIMPLEXML} corresponds to a different kind in a way that is indicated by the list item headers.
\begin{itemize}
	\item \textbf{ANTLR v4 keywords} are required by the notation.
	\item \antlrparserrule{\textbf{Parser rules}} describe the structure of a language.
		Alternatives on the right side of a rule are separated by the pipe character ($|$).
	\item \antlrlexerrule{\textbf{Lexer rules}} describe terminal symbols that the parser matches against the input string.
		A terminal symbol can be encoded as a string value or using a regular expression.
		The \textbf{fragment} keyword states that the lexer rule just brings more clarity into the notation but it is not visible in the parser output.
	\item \antlrliteralnoap{\textbf{String literals}}, always written inside of a pair of single quote marks, also represent terminal symbols, but by exact match to a string constant.
	\item \antlrregex{\textbf{Regular expressions}} describe string tokens to be matched against regular expressions with a special ANTLR v4 regex notation\footnote{https://github.com/antlr/antlr4/blob/master/doc/lexer-rules.md}.
\end{itemize}
Elements on the right side of a rule can be annotated with standard EBNF operators (\code{?}, \code{+}, \code{*}) that specify the allowed number of occurrences.

\subsection{Parsing Input Grammar}

The first phase in the process of MPS language construction is parsing of the input grammar that is defined using the ANTLR v4 notation.
An output of this phase is an AST representing the grammar.
Nodes of the AST correspond to rules and other grammar elements.

Nevertheless, the full AST, which comes out of the automatically generated parser, is quite complex and contains information not relevant for the INGRID algorithm.
In order to get a simple representation that is easy process by the later phases and keeps only information necessary for the construction of MPS languages, several steps of post-processing of the AST have to be performed in this phase.

A simplified tree of objects is derived from the AST in two passes over the grammar.
The result of the first pass is a tree where each node contains names of other referenced rules in the form of strings.
In the second pass, the whole tree is traversed and all the references are resolved to have a form of real pointers to other rule objects.

The representation of lexer rules (tokens) in the AST has to be simplified too.
Note that, in the ANTLR notation, lexer rules can be built from alternatives just like the parser rules --- see, for example, the lexer rule \antlrlexerrule{Name} from our SimpleXML language (Figure~\ref{fig:SIMPLEXML}).
For each lexer rule, the parser produces a tree that captures the structure of the rule.
However, the lexer rules are, in fact, just regular expressions used to recognize tokens in the input string.

Every tree that represents a lexer rule is flattened into the equivalent regular expression by the recursive algorithm specified in Figure~\ref{fig:ALGFLATTEN}.
A distinct sequence is created from the elements of each alternative, and then all those sequences are joined by the alternation operator \code{$|$} to form a regular expression.

\begin{figure}[ht]
\centering
\begin{framed}
\begin{alltt}
Flatten(\textit{R}):
  \textit{T} = empty list
  \textbf{for each} alternative \textit{A} \textbf{in} rule \textit{R}:
    \textit{R} = \antlrap\antlrap
    \textbf{for each} element \textit{E} \textbf{of} \textit{A}:
      \textbf{if} \textit{E} is not yet flattened \textbf{then}:
        Flatten(\textit{E})
      \textit{R}.append(\textit{E})
      \textit{R}.append(\textit{E}.operator)
    \textit{T}.add(\textit{R})
  build string S from elements of \textit{T}:
    \textit{S} = \(t\sb{1}\) | \(t\sb{2}\) | \(\ldots\) | \(t\sb{n}\)
  \textbf{return} \textit{S}
\end{alltt}
\end{framed}
\caption{Flattening algorithm}
\label{fig:ALGFLATTEN}
\end{figure}

The output of \texttt{Flatten(Name)}, i.e. application of the algorithm to the \antlrlexerrule{Name} rule, is this regular expression:

\begin{center}
  \texttt{[:a-zA-Z]([:a-zA-Z]|\textbackslash-|{\_}|\textbackslash.|[0-9])*}
\end{center}
It defines the syntactically valid identifiers of elements and attributes in SimpleXML.
Each identifier must begin with a letter, followed by any combination of letters, digits, underscore, dash, or a dot.

@@
